---
title: "geneSPIDER tutorial"
author: "Mateusz Garbulowski"
date: "2/2/2022"
output:
  html_document: default
  pdf_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Running geneSPIDER with synthetic expression data

This is a step-by-step tutorial of inferring a Gene Regulatory Network (GRN) with geneSPIDER for synthetic expression data
<center> 
```{r echo=FALSE, results='asis', warning=FALSE, fig.cap="Figure 1. Workflow of the geneSPIDER network inference"}
library(DiagrammeR)

mermaid("
graph TD
A1(Expression Data) --> B(Synthetic Data) 
B(Synthetic Data) --> A1(Expression Data)
A2(Scale-free Network) --> B 
B --> C(Inferred Network)
",
width=400,
height=400)
```
</center>

### Read the data

```{r echo=FALSE, results='asis'}
library(knitr)
cn <- rep(c("G1","G2","G3","...","Gn"), 2)
rn <- c("G1","G2","G3","...","Gn")
insd <- diag(5)
insd[,4] <- c("...","...","...","...","...")
insd[4,] <- c("...","...","...","...","...")
insd <- replace(insd, insd == "0", "...")
#insd <- replace(insd, insd == "1", "-1")
tab1 <- cbind(insd,insd)
colnames(tab1) <- cn
rownames(tab1) <- rn
kable(tab1, caption="Table.1 Format of Y matrix for two replicates")
```

```
### Construct a scale-free network
```{matlab}
% define the number of nodes based on the number of perturbed genes
N = siz(1);
% define desired sparsity degree, e.g. 3 edges per node
S = 3/(N); % please note, in geneSPIDER article this is N^2

% create scale-free network that gives probability to edges
A = datastruct.scalefree(N, S, randn(N) < S); % A is the network matrix
% set pin and pout parameters in order to remove selfloops
A = datastruct.cutSym(A, 0.6, 0.398); % probability of feedback loop is 1-(0.6+0.398)=0.002
% stabilize networks
A = datastruct.stabilize(A,'iaa','high');
% random weights cannot guarantee that the network is stable
% hence we stabilize, i.e. tune IAA degree

% create Network object
Net = datastruct.Network(A, 'yourNetwork');
```

### Construct a dataset

```{r echo=FALSE, results='asis'}
library(knitr)
cn <- rep(c("G1","G2","G3","...","Gn"), 2)
rn <- c("G1","G2","G3","...","Gn")
insd <- diag(5)
insd[,4] <- c("...","...","...","...","...")
insd[4,] <- c("...","...","...","...","...")
#insd <- replace(insd, insd == "0", "...")
tab1 <- cbind(insd,insd)
colnames(tab1) <- cn
rownames(tab1) <- rn
kable(tab1, caption="Table.2 Format of P matrix for two replicates")
```

```{matlab}
% define perturbation matrix for the experiment
P = [eye(N) eye(N)]; %sample perturbation matrix
% create data
X = Net.G*P; % G - static gain model, A - network

% create noise to data
% define signal-to-noise ratio
SNR = 0.1;
s = svd(X);
stdE = s(N)/(SNR*sqrt(chi2inv(1-analyse.Data.alpha,prod(size(P)))));
% estimate noise matrix
E = stdE*randn(size(P));
% input noise matrix
F = zeros(size(P));

Y = X+E; %response matrix + noise

% assign scale-free network
D(1).network = Net.network;
% define zero noise
D(1).E = E;
D(1).F = F;
D(1).Y = Y;
D(1).P = P;
D(1).lambda = [stdE^2,0];
D(1).cvY = D.lambda(1)*eye(N);
D(1).cvP = zeros(N);
D(1).sdY = stdE*ones(size(D.P));
D(1).sdP = zeros(size(D.P));

% create a data object with your data and scale-free network
Data = datastruct.Dataset(D,Net);
```

### Infer a network

```{r echo=FALSE, results='asis'}
library(knitr)
cn <- c("LSCO","LSCON","RNICO","GENIE3","LASSO","CLR")
rn <- c("Method","Function")
insd0 <- c("lsco",
          "LSCON",
          "RNI",
          "Genie3",
          "lasso",
          "CLR")
insd <- c("Least Squares Cut Off",
          "Least Squares Cut-Off with Normalization",
          "Robust Network Inference Cut Off",
          "GEne Network Inference with Ensemble of trees",
          "Least Absolute Shrinkage and Selection Operator",
          "Context Likelihood of Relatedness")
tab1 <- t(data.frame(insd, insd0))
colnames(tab1) <- cn
rownames(tab1) <- rn
kable(tab1, caption="Table.3 Some inference methods")
```

```{matlab}
zeta = logspace(-6,0,30);
infMethod = 'LSCON';
[Aest0, z0] = Methods.(infMethod)(Data,zeta);

% compare models
M = analyse.CompareModels(Net,Aest0);
M.AUROC
```

```{r echo=FALSE, results='asis'}
library(knitr)
cn <- c("G1","G2","G3","...","Gn")
rn <- c("G1","G2","G3","...","Gn")
insd <- diag(5)
insd[,4] <- c("...","...","...","...","...")
insd[4,] <- c("...","...","...","...","...")
insd <- replace(insd, insd == "0", "...")
tab1 <- insd
colnames(tab1) <- cn
rownames(tab1) <- rn
kable(tab1, caption="Table.4 Inferred network. Must be read from rows to columns.")
```